---
title: "PAIBI"
output: html_notebook
editor_options: 
chunk_output_type: inline
---

```{r setup}
if(!require(knitr)){install.packages("knitr", dependencies = TRUE)}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "C:/Users/West Liberty U/Desktop/Garrett_Hoover/Garrett_Hoover_thesis_documents")
getwd()
setwd("C:/Users/West Liberty U/Desktop/Garrett_Hoover/Garrett_Hoover_thesis_documents")
```
```{r}
if(!require(dplyr)){install.packages("dplyr", dependencies = TRUE)}
if(!require(tidyr)){install.packages("tidyr", dependencies = TRUE)}
if(!require(readxl)){install.packages("readxl", dependencies = TRUE)}
if(!require(writexl)){install.packages("writexl", dependencies = TRUE)}
if(!require(conflicted)){install.packages("conflicted", dependencies = TRUE)}

library ("conflicted")
library ("dplyr")
library("tidyr")
library("readxl")
library("writexl")

conflict_prefer("filter", "dplyr")

sessionInfo()

#Designated Use Codes
use <-data.frame(INT_USE_GEN_ID=c(1:8, 21), 
                    STR_USE_CODE=c("CWF", "EV", "HQ", "HQ-CWF", "HQ-TSF", "HQ-WWF", "TSF", "WWF", "N/A"),
                    STR_USE_DESCRIPTION=c("CWF(COLD WATER FISHES)", "EV(EXCEPTIONAL VALUE)", "HQ(HIGH QUALITY)", "HQ-CWF(HIGH QUALITY-COLD WATER FISHES", "HQ-TSF(HIGH QUALITY-TROUT STOCKING)", "HQ-WWF(HIGH QUALITY-WARM WATER
                                          FISHES)", "TSF(TROUT STOCKING)", "WWF(WARM WATER FISHES)", "Missing from Chapter 93"))

use<- read_excel("SLIMS_DBA_MI_BUG_CODES_LUT.XLSX",2)

#Taxa Table
lut <- read_excel("SLIMS_DBA_MI_BUG_CODES_LUT.XLSX",1)
```
This code was written by Lou Reynolds and Leah Ettema - USEPA
reynolds.louis@epa.gov  ettema.leah@epa.gov
Contact us with problems or questions.

Make an attribute table to compute metrics.
The basis of this is the "SLIMS_DBA_MI_BUG_CODES_LUT.XLSX"
This table was supplied by PADEP in Feb 2018.

```{r Construct Attribute Table}
str(lut)

conflict_prefer("select", "stats")
#These lines create the basis for any metrics one might want based on traits 
rtax <- lut %>% 
  mutate(hilsid0=ifelse(INT_HILS_ID_LEVEL==0,1,0),
         hilsid1=ifelse(INT_HILS_ID_LEVEL==1,1,0),
         hilsid2=ifelse(INT_HILS_ID_LEVEL==2,1,0),
         hilsid3=ifelse(INT_HILS_ID_LEVEL==3,1,0),
         hilsid4=ifelse(INT_HILS_ID_LEVEL==4,1,0),
         hilsid5=ifelse(INT_HILS_ID_LEVEL==5,1,0),
         hilsid6=ifelse(INT_HILS_ID_LEVEL==6,1,0),
         hilsid7=ifelse(INT_HILS_ID_LEVEL==7,1,0),
         hilsid8=ifelse(INT_HILS_ID_LEVEL==8,1,0),
         hilsid9=ifelse(INT_HILS_ID_LEVEL==9,1,0),
         hilsid10=ifelse(INT_HILS_ID_LEVEL==10,1,0),
         #sensitive ept
         e=ifelse(INT_BUG_CODE >= 1020000000&INT_BUG_CODE <= 1029999999&INT_HILS_ID_LEVEL<=4,1,0),
         p=ifelse(INT_BUG_CODE >= 1040000000&INT_BUG_CODE <= 1049999999&INT_HILS_ID_LEVEL<=4,1,0),
         t=ifelse(INT_BUG_CODE >= 1080000000&INT_BUG_CODE <= 1089999999&INT_HILS_ID_LEVEL<=4,1,0),
         ept=ifelse(e+p+t>0,1,0), #all sensitive ept
         sens=ifelse(INT_HILS_ID_LEVEL<=3,1,0),#sesitive taxa
         bcg123sm=ifelse(INT_BCG_COLD<=3&INT_BCG_COLD>0,1,0),
         bcg456sm=ifelse(INT_BCG_COLD>=4,1,0),
         bcg123lg=ifelse(INT_BCG_WARM<=3&INT_BCG_WARM>0,1,0),
         bcg456lg=ifelse(INT_BCG_WARM>=4,1,0),
         eall = ifelse(INT_BUG_CODE >= 1020000000&INT_BUG_CODE <= 1029999999,1,0),
         pall = ifelse(INT_BUG_CODE >= 1040000000&INT_BUG_CODE <= 1049999999,1,0),
         tall = ifelse(INT_BUG_CODE >= 1080000000&INT_BUG_CODE <= 1089999999,1,0), #ALL EPT
         eptall = ifelse(eall+pall+tall>0,1,0),
         mayfly = ifelse(eall==1,1,0), #Mayfly Present/Absent
         leuctra_amphi = ifelse(STR_TAXA_NAME=="Leuctra"|STR_TAXA_NAME=="Amphinemura",1,0))

#Put rtax into "long" or stacked form - gather - to compute totals
rtaxl <- rtax %>% 
  dplyr::select(STR_TAXA_NAME,hilsid0,hilsid1,hilsid2,hilsid3,hilsid4,hilsid5,hilsid6,hilsid7,hilsid8,hilsid9,hilsid10,ept,sens,bcg123sm,bcg456sm,bcg123lg,bcg456lg,eall,pall,tall,eptall,mayfly,leuctra_amphi) %>% 
  filter(!is.na(STR_TAXA_NAME)) %>% 
  gather(key=metric, value=metric_val, -STR_TAXA_NAME) %>% 
  filter(metric_val>0) #%>% 
  #distinct() %>% #Gets rid of all duplicates such as Chironomidae
  #filter(STR_TAXA_NAME=="Chironomidae")

```

Input Data
This R code assumes that bug data is in the form of 
SampleID  taxaname  COUNT
The taxa name should be renamed to STR_TAXA_NAME

There also needs to be separate Station Info data

SampleID month WSA  INT_USE_GEN_ID  BaselineIBI 

SampleID=common field with the bug data
month=numerical month that sample collected (1-12)
WSA=watershed area in Square Miles
INT_USE_GEN_ID=Use ID as in table use
BaselineIBI=past IBI if there is one.  NA if not

The raw data may not have met the total number of individuals requirements of the PAIBI.  A subsampling routine was added in this event for situations where the total number of individuals is >240.

  NOTE: This is dummied data - it only exists as an example and to test the code.

To obtain month/year from a consistently formated sample ID, use something like: 
  separate(SampleID, into=c("year","rest"), sep=2,remove=FALSE) %>% 
  separate(rest, into=c("month","rest1"), sep=2,remove=FALSE) %>%
  select(-rest, -rest1) 

```{r Input data}

#Read in Raw data
bugdat <- read_excel("bugs_year2.xlsx",sheet="samples")
bugsta <- read_excel("bugs_year2.xlsx",sheet="stations")
      
# This little bit accounts for multiple STR_ID's in a sample
#- e.g. Lumbriculidae, etc.
bugdat <-bugdat %>% 
  group_by(SampleID,STR_TAXA_NAME) %>% 
  summarise (totn = sum(Count)) %>% 
  ungroup()

#This removes ID's only to Diptera
bugdat<-bugdat %>%
  filter(STR_TAXA_NAME!="Diptera")

#Subsample if needed--------

#Total number of individuals per sample
bugtotalsraw <- bugdat %>% 
  group_by(SampleID) %>% 
  summarise(bugtotalraw=sum(totn)) %>%  
  dplyr::select(SampleID, bugtotalraw) #bugtotal is total

#Table displaying sampleID's with unusual counts (outside of 160-240).
SampleSizeWarning<- bugtotalsraw %>% 
  filter(bugtotalraw>240 | bugtotalraw < 160)

#When bug total is greater than 240, random sample down to 200
LargeSample<-SampleSizeWarning %>%
  filter(bugtotalraw>240) 

#LS=Large Sample
LS <- left_join(LargeSample,bugdat,by="SampleID")  
LSrow<-nrow(LS)

for(s in 1:LSrow) #this loop copies and pastes a new row totn-1 number of times, making a unique row for each individual taxa
{
  if (LS$totn[s]>1)
  {
    reps=LS$totn[s]-1 #number of individuals per STR_TAXA_NAME -1
    for (t in 1:reps)
    {
      LS<-rbind(LS, LS[s,]) #paste the row "reps" number of times
    }
  }
} #totn is now meaningless (should be 1 for each row), but will delete in next select()
as.tbl(LS) #need this for sample_n function
LSnew <- LS %>%
  group_by(SampleID) %>%
  dplyr::select(-totn, -bugtotalraw)  %>%
  sample_n(240, replace=FALSE) #randomly selects 240 rows (individuals) for each SampleID

LSnewsum1<-LSnew %>%  #recounting (summarizing and collapsing) the number of each taxa per sampleID
  group_by(SampleID, STR_TAXA_NAME) %>%
  summarise(totn=n()) %>%
  mutate(subsamp=as.character("Y")) #These SampleID's were subsampled

#Making bugdat table for analysis, with subsampled data and column indicating it was subsampled 
samplematch<-bugdat$SampleID  %in%  LSnewsum1$SampleID #creating T/F vector: if Subsampled Sample ID's = Sample ID's in bugdat, True means SampleID was subsampled

bugdatn <- bugdat %>% #making columns match for bind
  mutate(subsamp=as.character("N")) #indicating not subsampled

bugdatn<-bugdatn[!samplematch,] #delete rows that were subsampled (select only rows that do not have matches with subsampled SampleIDs)
bugdat<-bind_rows(bugdatn, LSnewsum1) #merge rows with the new counts that were subsampled

subsamp<-bugdat %>%
  select(SampleID, subsamp) 
#will paste on to IBI score Final
subsamp<-unique(subsamp) 
################end subsampling #########

```

The following are all calculations for PAIBI metrics

```{r PAIBI calculations}
#Total number of individuals per sample
bugtotals <- bugdat %>% 
  group_by(SampleID) %>% 
  summarise(bugtotal=sum(totn)) %>% 
  select(SampleID, bugtotal)

#Put the bug data together with the Traits

bugxtax <- left_join(bugdat,rtaxl,by="STR_TAXA_NAME") 

#a few QA checks  Feel free to add more...
QA_bySample <- bugxtax %>% 
  left_join(bugtotals, by="SampleID") %>% 
   arrange(SampleID, metric)

QA_nannies<- bugxtax %>%  ###check of missing values
  filter(is.na(metric))

bugxtax <- bugxtax %>%  ##Remove metrics with NA values
  filter(!is.na(metric))
names(bugxtax)

#bug names that are in the sample but not the LUT
missingbugs<-bugdat$STR_TAXA_NAME %in% lut$STR_TAXA_NAME
missingbugs<-bugdat[missingbugs==FALSE,]

write_xlsx(list(SampleMetricQANA=QA_nannies, SampleSizeWarning=SampleSizeWarning, BugsnotinLUT=missingbugs), "Bugdat_QA.xlsx")

#Preparing for metric calculations
 #Metrich = taxa richness of each metric by sample
 #Metsum = number of individuals per metric by sample 
met1 <- bugxtax %>%
  group_by(SampleID, metric) %>% 
  summarise(Metrich=length(totn), Metsum=sum(totn)) %>% 
  ungroup()

 #Metprop=proprotion of individuals in the sample with that metric  
met2 <- left_join (met1, bugtotals, by="SampleID") %>% 
  mutate(Metprop=Metsum/bugtotal)

#The following are the IBI metrics 

#Total Richness
rich <- bugdat %>% 
  group_by(SampleID) %>% 
  summarise(Richness=n_distinct(STR_TAXA_NAME))

#Sensitive EPT
eptrich <- met2 %>%
  filter(metric=="ept") %>% 
  group_by(SampleID) %>% 
  summarise(eptR=Metrich)#  changed from sum(Metrich) (made no difference really)

#To calculate Becks Index, the metrics need to be put into wide form - spread
becks <- met2 %>%
  select(-Metsum,-bugtotal,-Metprop) %>% 
  spread(metric,Metrich) 

becks[is.na(becks)] <- 0 #Replace NANs in table with zeros

becks <- becks %>% 
  mutate(becksindex=(3*hilsid0+2*hilsid1+1*hilsid2)) %>% 
  select(SampleID,becksindex)

#Hisenhoff Index

#Hilsenhoff Index taxa info
hiltax <- rtax %>% 
  select(STR_TAXA_NAME, INT_HILS_ID_LEVEL) %>% 
  ##distinct() %>% #Gets rid of all duplicates such as Chironomidae
  filter(!is.na(INT_HILS_ID_LEVEL))# Gets rid of NA's in list

#Raw Calculations for Hilsenhoff Index and Shannon Diversity  
hils <- bugdat %>% 
  left_join(hiltax) %>% 
  group_by(SampleID)%>% 
  mutate(nxa=totn*INT_HILS_ID_LEVEL, sumn=sum(totn))
  
#Hilsenhoff Index
hilsindx <- hils %>% 
  group_by(SampleID) %>% 
  summarise(hilsindex=sum(nxa)/mean(sumn))
  
#Shannon Diversity
shandiv <- hils %>%
  mutate(pi=totn/sumn,lnpi=log(pi),pilnpi=pi*lnpi) %>% 
  group_by(SampleID) %>% 
  summarise(shdv=-sum(pilnpi))

#Percent Sensitive Individuals
psens <- met2 %>%
  select(SampleID, metric, Metprop) %>% 
  filter(metric=="sens") %>% 
  mutate(psen=Metprop*100) %>% 
  select(SampleID, psen)

#all ept
allept <- met2 %>% 
  filter (metric=="eptall") %>% 
  group_by(SampleID) %>% 
  mutate(allept=Metrich)

#Put all the tables together
ibi_1 <- left_join(rich,eptrich) %>%  
  left_join(becks) %>% 
  left_join(hilsindx) %>% 
  left_join(shandiv) %>% 
  left_join(psens)
  
ibi_1$psen<-replace(ibi_1$psen, which(is.na(ibi_1$psen)), as.numeric(0)) #if there are no sensitive taxa, change NA to 0
ibi_1$eptR<-replace(ibi_1$eptR, which(is.na(ibi_1$eptR)), as.numeric(0)) #if there are no ept taxa, change NA to 0

ibi <- left_join(ibi_1,bugsta, by = "SampleID") %>% 
  #left_join (gates, by="SampleID") %>% 
  left_join (use, by="INT_USE_GEN_ID")

ibi$season <- ifelse(ibi$month >=6& ibi$month <= 9, "summer",
                       ifelse(ibi$month>=11| ibi$month <=5,"winter",
                              ifelse(ibi$month ==10,"bpj","NA")))
ibi$size <- ifelse(ibi$WSA <=25,"small", "large")

###IBI is scored to different standard values according to Watershed Area.
#small is <25 square miles and large is >50 square miles.
#PADEP scores them both and the biologists decide which to apply to streams between 25-50.

#This function truncates scores to fall between 0-100
truncfxn <- function(x){ifelse(x>=100,100,x)}
#compute IBI scores
ibiscore_final <- ibi %>%
  mutate(sc_sm_Richness=truncfxn((Richness/33)*100),
         sc_sm_eptR=truncfxn((eptR/19)*100),
         sc_sm_becks=truncfxn((becksindex/38)*100),
         sc_sm_hils=truncfxn(((10-hilsindex)/(10-1.89))*100),
         sc_sm_shdv=truncfxn((shdv/2.86)*100),
         sc_sm_psen=truncfxn((psen/84.5)*100),
         ibi_sm=(sc_sm_Richness+sc_sm_eptR+sc_sm_becks+sc_sm_hils+sc_sm_shdv+sc_sm_psen)/6) %>% 
  mutate(sc_lg_Richness=truncfxn((Richness/31)*100),
         sc_lg_eptR=truncfxn((eptR/16)*100),
         sc_lg_becks=truncfxn((becksindex/22)*100),
         sc_lg_hils=truncfxn(((10-hilsindex)/(10-3.05))*100),
         sc_lg_shdv=truncfxn((shdv/2.86)*100),
         sc_lg_psen=truncfxn((psen/66.7)*100),
         ibi_lg=(sc_lg_Richness+sc_lg_eptR+sc_lg_becks+sc_lg_hils+sc_lg_shdv+sc_lg_psen)/6) %>% 
  mutate(BaselineIBI=(ifelse(is.na(BaselineIBI),ibi_sm,BaselineIBI))) #Baseline IBI score added if it exists, if not, small stream IBI score used

#add subsamp column to IBIscore_final
ibiscore_final <- full_join(ibiscore_final, subsamp, by="SampleID")

names(ibiscore_final)
``` 

The following is preparation for the gates (othgate calculations below). 

Gates are questions used in the PADEP flowchart to determine impairment.

Following calculation of the gate endpoints, attainment decisions can be made.  

```{R Gates from the Flow Chart}
#Gate 1
#Are mayflies, stoneflies, or caddislfies absent from the sub-sample?
ept_pa <- met2 %>% 
  filter(metric%in%c("eall", "pall","tall")) %>% 
  select(SampleID,metric,Metrich) %>% 
  spread(metric,Metrich) %>% 
  mutate(eall=replace(eall,which(is.na(eall)),0),
         pall=replace(pall,which(is.na(pall)),0),
         tall=replace(tall,which(is.na(tall)),0),
         ept_pa = ifelse(eall>=1&pall>=1&tall>=1,1,0)) %>% 
  select(SampleID, ept_pa)

#Gate 3 
#Is the ratio of BCG attribute 1,2,3 taxa to BCG attribute 4,5,6 taxa<0.75 

bcg_rich <- met2 %>% #This calculates the bcg richness ratios
  select(SampleID, metric, Metrich) %>% 
  filter(metric %in% c("bcg123sm","bcg456sm","bcg123lg","bcg456lg")) %>% 
  spread(metric,Metrich) %>% 
  mutate(bcgr_sm=bcg123sm/bcg456sm,
         bcgr_lg=bcg123lg/bcg456lg) %>% 
  select(SampleID, bcgr_sm,bcgr_lg)

bcg_ind <- met2 %>% #This calculates the bcg individuals ratios
  select(SampleID, metric, Metsum) %>% 
  filter(metric %in% c("bcg123sm","bcg456sm","bcg123lg","bcg456lg")) %>% 
  spread(metric,Metsum) %>% 
  mutate(bcgi_sm=bcg123sm/bcg456sm,
         bcgi_lg=bcg123lg/bcg456lg) %>% 
  select(SampleID, bcgi_sm,bcgi_lg)

#Gate 4
#Does the sub-sample show signatures of acidification year-round.  (i.e. <3 mayfly taxa)
# with mayflies representing <5% and Leuctra+Amphinemura representing >25% of individuals in the sub-sample?
acid1 <- met2 %>% 
  select(SampleID,metric,Metrich,Metprop) %>% 
  filter(metric %in% c("mayfly")) %>% 
  mutate (mayflyrich=Metrich, mayflypr=Metprop) %>% 
  select (-metric,-Metrich,-Metprop)

acid2 <- met2 %>% 
  select (SampleID,metric,Metprop) %>% 
  filter(metric %in% c("leuctra_amphi")) %>% 
  group_by(SampleID) %>% 
  summarise(leucamph_pa=sum(Metprop)) 
  

gates <- left_join(ept_pa,bcg_ind) %>% 
  left_join(bcg_rich) %>% 
  left_join(acid1) %>% 
  left_join(acid2) %>% 
  mutate(leucamph_pa=replace(leucamph_pa,which(is.na(leucamph_pa)),0)) 

#Attainment of EV and HQ Waters
ibiscore_final <- ibiscore_final %>% #Impairment decisions for Exceptional Value and High Quality waters
mutate(hqevcut_sm = ibi_sm-BaselineIBI,
       hqevcut_lg = ibi_lg-BaselineIBI,
       #creating 3 columns based on stream size and season; user decides which is appropriate
       hqevw_wsm_attdec = ifelse(ibi_sm >=63 & hqevcut_sm <=11, "attains", 
                                "impaired"), #winter small stream (wsm)
       hqevw_wlg_attdec = ifelse(ibi_sm >=63 & hqevcut_lg <=11, "attains", 
                                "impaired"), #winter large stream (wlg)
       hqevw_s_attdec = "resample") #any sized stream in summer

#Attainment decisions for other use waters
ibiscore_final <- ibiscore_final %>%
  left_join (gates, by="SampleID") %>%
  #assigned 1 if attaining, 0 if not
  mutate(othgate_ept_lg=ifelse(ept_pa>=1,1,0),
         othgate_becksen_lg=ifelse(sc_lg_becks>=33.3 & sc_lg_psen>=25, 1,0),
         othgate_bcg_lg=ifelse(bcgr_lg>=0.75 & bcgi_lg>=0.75, 1, 0),
         othgate_acid_lg=ifelse(mayflyrich>3 & mayflypr>0.05 & leucamph_pa<0.25,1,0),
         othgateall_lg=(othgate_ept_lg+othgate_becksen_lg+othgate_bcg_lg+othgate_acid_lg)) %>% 
  mutate(othgate_ept_sm=ifelse(ept_pa>=1,1,0),
         othgate_becksen_sm=ifelse(sc_sm_becks>=33.3 & sc_sm_psen>=25, 1,0),
         othgate_bcg_sm=ifelse(bcgr_sm>=0.75 & bcgi_sm>=0.75, 1, 0),
         othgate_acid_sm=ifelse(mayflyrich>3 & mayflypr>=0.05 & leucamph_pa<0.25,1,0),
         othgateall_sm=(othgate_ept_sm+othgate_becksen_sm+othgate_bcg_sm+othgate_acid_sm))

#if the IBI score passes and all 4 gates, the site attains
ibiscore_final <- ibiscore_final %>% 
mutate(othwlg_attdec=ifelse(ibi_lg>=50 & othgateall_lg==4, "attains", "impaired"), #winter large
       othslg_attdec=ifelse(ibi_lg>=43 & othgateall_lg==4, "attains", "impaired"), #summer large
       othwsm_attdec=ifelse(ibi_sm>=50 & othgateall_sm==4, "attains", "impaired"), #winter small
       othssm_attdec=ifelse(ibi_sm>=43 & othgateall_sm==4, "attains", "impaired")) #summer small

#output attainment decision points for all contingencies
attains <- ibiscore_final %>% 
  select(SampleID,STR_USE_CODE,WSA,season, size, ibi_sm, ibi_lg, hqevw_wsm_attdec,hqevw_wlg_attdec, hqevw_s_attdec,othwlg_attdec,othslg_attdec,
         othwsm_attdec,othssm_attdec)

#Output to XLSX.  
#Modify code to output any table in work env. to a worksheet 
#list(worksheetname=tablname,...)
write_xlsx(list(paibi_sc=ibiscore_final, pa_ibi_attain=attains),"PAibi.xlsx") 

```

