---
title: "R Notebook"
output: html_notebook
---

```{r}
if(!require("installr")){install.packages("installr", dependencies = TRUE)}
library(installr)
updateR()
if(!require("INLA")){install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/testing"), dep=TRUE)}
library(INLA)
```

```{r setup}
if(!require(knitr)){install.packages("knitr", dependencies = TRUE)}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "C:/Users/West Liberty U/Desktop/Garrett_Hoover/Garrett_Hoover_thesis_documents")
getwd()
setwd("C:/Users/West Liberty U/Desktop/Garrett_Hoover/Garrett_Hoover_thesis_documents")
```
```{r}
#Install what you need. I don't remember all the packages necessary to run the code, but if you run all of these packages
#the code should run

if(!require(nlme)){install.packages("nlme", dependencies = TRUE)}
library(nlme)
if(!require(MASS)){install.packages("MASS", dependencies = TRUE)}
library(MASS)
if(!require(MuMIn)){install.packages("MuMIn", dependencies = TRUE)}
library(MuMIn)
if(!require(lme4)){install.packages("lme4", dependencies = TRUE)}
library(lme4)
if(!require(ggplot2)){install.packages("ggplot2", dependencies = TRUE)}
library(ggplot2)
if(!require(dplyr)){install.packages("dplyr", dependencies = TRUE)}
library(dplyr)
if(!require(psych)){install.packages("psych", dependencies = TRUE)}
library(psych)
if(!require(usdm)){install.packages("usdm", dependencies = TRUE)}
library(usdm)
if(!require(lattice)){install.packages("lattice", dependencies = TRUE)}
library(lattice)
if(!require(car)){install.packages("car", dependencies = TRUE)}
library(car)
if(!require(MCMCglmm)){install.packages("MCMCglmm", dependencies = TRUE)}
library(MCMCglmm)
if(!require(vegan)){install.packages("vegan", dependencies = TRUE)}
library(vegan)
if(!require(grid)){install.packages("grid", dependencies = TRUE)}
library(grid)
if(!require(ggrepel)){install.packages("ggrepel", dependencies = TRUE)}
library(ggrepel)
if(!require(ggpubr)){install.packages("ggpubr", dependencies = TRUE)}
library(ggpubr)
if(!require(INLA)){install.packages("INLA", dependencies = TRUE)}
library(INLA)
if(!require(ggregplot)) devtools::install_github("gfalbery/ggregplot")
library(ggregplot)
if(!require(tidyverse)){install.packages("tidyverse", dependencies = TRUE)}
library(tidyverse)
if(!require(snow)){install.packages("snow", dependencies = TRUE)}
library(snow)
if(!require(parallel)){install.packages("parallel", dependencies = TRUE)}
library(parallel)
if(!require(landscapemetrics)){install.packages("landscapemetrics", dependencies = TRUE)}
library(landscapemetrics)
if(!require(rgdal)){install.packages("rgdal", dependencies = TRUE)}
library(rgdal)
#if(!require(GARPTools)){install.packages("GARPTools", dependencies = TRUE)}
#library(GARPTools)
if(!require(maps)){install.packages("maps", dependencies = TRUE)}
library(maps)
if(!require(sp)){install.packages("sp", dependencies = TRUE)}
library(sp)
if(!require(rgeos)){install.packages("rgeos", dependencies = TRUE)}
library(rgeos)
if(!require(maptools)){install.packages("maptools", dependencies = TRUE)}
library(maptools)
if(!require(rgeos)){install.packages("rgeos", dependencies = TRUE)}
library(rgeos)
if(!require(mapdata)){install.packages("mapdata", dependencies = TRUE)}
library(mapdata)
if(!require(gbm)){install.packages("gbm", dependencies = TRUE)}
library(gbm)
if(!require(dismo)){install.packages("dismo", dependencies = TRUE)}
library(dismo)
if(!require(ResourceSelection)){install.packages("ResourceSelection", dependencies = TRUE)}
library(ResourceSelection)
if(!require(SDMPlay)){install.packages("SDMPlay", dependencies = TRUE)}
library(SDMPlay)
if(!require(spatstat)){install.packages("spatstat", dependencies = TRUE)}
library(spatstat)
if(!require(openxlsx)){install.packages("openxlsx", dependencies = TRUE)}
library(openxlsx)
if(!require(spatialEco)){install.packages("spatialEco", dependencies = TRUE)}
library(spatialEco)
if(!require(pROC)){install.packages("pROC", dependencies = TRUE)}
library(pROC)
if(!require(randomForest)){install.packages("randomForest", dependencies = TRUE)}
library(randomForest)
if(!require(boot)){install.packages("boot", dependencies = TRUE)}
library(boot)
if(!require(Hmisc)){install.packages("Hmisc", dependencies = TRUE)}
library(Hmisc)
if(!require(verification)){install.packages("verification", dependencies = TRUE)}
library(verification)
if(!require(pdp)){install.packages("pdp", dependencies = TRUE)}
library(pdp)
if(!require(caret)){install.packages("caret", dependencies = TRUE)}
library(caret)
if(!require(ROCR)){install.packages("ROCR", dependencies = TRUE)}
library(ROCR)
#if(!require(TSS)){install.packages("TSS", dependencies = TRUE)}
#library(TSS)
if(!require(openxlsx)){install.packages("openxlsx", dependencies = TRUE)}
library(openxlsx)
if(!require(landscapetools)){install.packages("landscapetools", dependencies = TRUE)}
library(landscapetools)
if(!require(sf)){install.packages("sf", dependencies = TRUE)}
library(sf)
if(!require(purrr)){install.packages("purrr", dependencies = TRUE)}
library(purrr)
if(!require(reshape2)){install.packages("reshape2", dependencies = TRUE)}
library(reshape2)
if(!require(lmPerm)){install.packages("lmPerm", dependencies = TRUE)}
library(lmPerm)
#if(!require(rstatix)){install.packages("rstatix", dependencies = TRUE)}
#library(rstatix)
if(!require(spdep)){install.packages("spdep", dependencies = TRUE)}
library(spdep)
#if(!require(SDMTools)){install.packages("SDMTools", dependencies = TRUE)}
#library(SDMTools)
```
```{r}
install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
remotes::install_github("gfalbery/ggregplot")
install.packages("R.utils")
install.packages("SDMTools","http://rforge.net/",type="source")
install.packages("remotes")
remotes::install_github("cghaase/GARPTools")

packages = c("nlme","MASS", "MuMIn", "lme4", "usdm", "ggplot2", "lattice", "car", "MCMCglmm",
             "vegan", "grid", "ggrepel", "ggpubr", "snow", "parallel", "landscapemetrics",
             "raster", "rgdal", "GARPTools", "maps", "sp", "rgeos", "maptools", "mapdata",
             "gbm", "dismo", "ResourceSelection", "dplyr", "tidyverse", "SDMPlay", "spatstat", "xlsx",
             "spatialEco", "pROC", "randomForest", "boot", "Hmisc", "verification", "pdp", "caret", "ROCR", "TSS",
             "openxlsx", "landscapetools", "sf", "purrr", "reshape2", "rstatix", "lmPerm", "INLA", "ggregplot", "spdep")
```

## Now load or install&load all
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

#Call in the data------
lm <- read.csv("C:/Users/West Liberty U/Desktop/Garrett_Hoover/Garrett_Hoover_thesis_documents/lm.csv")
na.omit(lm)
attach(lm)

#Multicollinearity------- 
test <- data.frame(scale(temp), scale(turb), scale(do), scale(ph), scale(cond), scale(orp), scale(cl), 
                      scale(percent_dev), scale(percent_forest), scale(percent_ag), scale(percent_dev_riparian), scale(percent_forest_riparian),
                      scale(percent_ag_riparian), scale(watershed_area), scale(ohqhei), scale(pawqnha))
test_1 <- vifstep(test, th = 10)
test_1

test_x <- data.frame(scale(temp), scale(turb), scale(do), scale(ph), scale(cond), scale(orp), scale(cl), 
                      scale(percent_dev), scale(percent_ag), scale(percent_dev_riparian),
                      scale(percent_ag_riparian), scale(watershed_area), scale(ohqhei), scale(pawqnha))
test_x1 <- vifstep(test_x, th = 10)
test_x1

#Produces Function for INLA models-------
f1 = ibi ~ scale(turb) + scale(do) + scale(ph) + scale(cond) + scale(orp) + scale(cl) + scale(percent_dev) +
scale(percent_ag) + scale(percent_dev_riparian) + scale(percent_ag_riparian) + scale(ohqhei) + f(site, model = "iid")

fx = ibi ~ scale(turb) + scale(do) + scale(ph) + scale(cond) + scale(orp) + scale(cl) + scale(percent_dev) +
scale(percent_ag) + scale(percent_dev_riparian) + scale(percent_ag_riparian) + scale(ohqhei)

inla_cul_model <- inla(f1, family = "gaussian", control.compute = list(dic = TRUE), data = lm)
summary(inla_cul_model)

inla_cul_model_x <- inla(fx, family = "gaussian", control.compute = list(dic = TRUE), data = lm)
summary(inla_cul_model_x)
#Produces caterpillar plot for parameter effect sizes (ggregplot) 
Efxplot(list(inla_cul_model))

#Variable reduction function I adjusted from the ggregplot package---------
INLAModelSel_sc <- function(Response, Explanatory, Random = NULL, RandomModel = NULL, Family, Data, Delta = 2){
  
  require(INLA); require(ggplot2)
  
  Explanatory2 <- paste(Explanatory, collapse = " + ")
  
  if(!is.null(Random)){
    Random2 <- paste(paste0("f(",Random, ", model = '", RandomModel, "')"), collapse = " + ")
    
    f1 <- as.formula(paste0(Response, " ~ ", paste(Explanatory2, " + ", Random2, collapse = " + ")))
  }else{
    f1 <- as.formula(paste0(Response, " ~ ", paste(Explanatory2, collapse = " + ")))
  }
  
  FullModel <-   inla(f1,
                      family = Family,
                      data = Data,
                      scale = TRUE,
                      verbose = TRUE,
                      control.compute = list(dic = TRUE))
  
  ModelList <- AllModelList <- RemovedList <- FullFormulaList <- FormulaList <- list()
  DICList <- dDICList <- list()
  
  DICList[[1]] <- FullModel$dic$dic
  FullFormulaList[[1]] <- f1
  
  
  for(x in 1:length(Explanatory)){
    
    Explanatory3 <- paste(Explanatory[-x], collapse = " + ")
    
    if(!is.null(Random)){
      
      f2 <- as.formula(paste0(Response, " ~ ", paste(Explanatory3, " + ", Random2, collapse = " + ")))
    } else{
      f2 <- as.formula(paste0(Response, " ~ ", paste(Explanatory3, collapse = " + ")))
      
    }
    
    Model1 <- inla(f2,
                   family = Family,
                   data = Data,
                   scale = TRUE,
                   verbose = TRUE,
                   control.compute = list(dic = TRUE))
    
    ModelList[[x]] <- Model1
    
    FormulaList[[x]] <- f2
    
  }
  
  DICList[[2]] <- sapply(ModelList, function(y) y$dic$dic)
  names(DICList[[2]]) <- Explanatory
  dDICList[[1]] <- DICList[[2]] - DICList[[1]]
  names(dDICList[[1]]) <- Explanatory
  RemovedList[[1]] <- Explanatory
  
  FullFormulaList[[2]] <- FormulaList
  
  AllModelList[[1]] <- FullModel
  AllModelList[[2]] <- ModelList
  
  NewExplanatory <- Explanatory
  
  while(min(dDICList[[length(dDICList)]]) < Delta&length(NewExplanatory)>0){
    
    print(paste("Losing", names(dDICList[[length(dDICList)]])[which(dDICList[[length(dDICList)]] == min(dDICList[[length(dDICList)]]))]))
    print(Text <- paste("Run", length(DICList)))
    
    ModelList <- FormulaList <- list()
    
    NewExplanatory <- NewExplanatory[-which(dDICList[[length(dDICList)]]==min(dDICList[[length(dDICList)]]))]
    
    if(length(NewExplanatory) > 0){
      for(x in 1:length(NewExplanatory)){
        
        Explanatory3 <- paste(NewExplanatory[-x], collapse = " + ")
        
        if(!is.null(Random)){
          
          f2 <- as.formula(paste0(Response, " ~ ", paste(Explanatory3, " + ", Random2, collapse = " + ")))
        } else{
          f2 <- as.formula(paste0(Response, " ~ ", paste(Explanatory3, collapse = " + ")))
          
        }
        
        Model1 <- inla(f2,
                       family = Family,
                       data = Data,
                       scale = TRUE,
                       verbose = TRUE,
                       control.compute = list(dic = TRUE))
        
        ModelList[[x]] <- Model1
        
        FormulaList[[x]] <- f2
        
      }
      
    }else{
      
      if(!is.null(Random)){
        
        f2 <- as.formula(paste0(Response, " ~ ", paste(Explanatory3, " + ", Random2, collapse = " + ")))
      } else{
        f2 <- as.formula(paste0(Response, " ~ ", paste(Explanatory3, collapse = " + ")))
        
      }
      
      Model1 <- inla(f2,
                     family = Family,
                     data = Data,
                     scale = TRUE,
                     verbose = TRUE,
                     control.compute = list(dic = TRUE))
      
      ModelList[[x]] <- Model1
      
      FormulaList[[x]] <- f2
      
    }
    
    AllModelList[[length(AllModelList)+1]] <- ModelList
    FullFormulaList[[length(FullFormulaList)+1]] <- FormulaList
    
    DICList[[length(DICList)+1]] <- sapply(ModelList, function(y) y$dic$dic)
    names(DICList[[length(DICList)]]) <- NewExplanatory
    dDICList[[length(dDICList)+1]] <- DICList[[length(DICList)]] - min(DICList[[length(DICList)-1]])
    names(dDICList[[length(dDICList)]]) <- NewExplanatory
    
    RemovedList[[length(RemovedList)+1]] <- NewExplanatory
    
  }
  
  FinalModel <- AllModelList[[length(AllModelList)-1]][[which(dDICList[[length(dDICList)-1]]==min(dDICList[[length(dDICList)-1]]))]]
  FinalFormula <- FullFormulaList[[length(AllModelList)-1]][[which(dDICList[[length(dDICList)-1]]==min(dDICList[[length(dDICList)-1]]))]]
  
  return(list(FinalModel = FinalModel,
              AllModels = AllModelList,
              Removed = RemovedList,
              DIC = DICList,
              dDIC = dDICList,
              FormulaList = FullFormulaList,
              FinalFormula = FinalFormula))
}

#All variables used in your first model
covar <- c("turb", "do", "ph", "cond", "orp", "cl", "percent_dev", "percent_ag", "percent_dev_riparian", "percent_ag_riparian", "ohqhei")

resp <- "ibi"

inla_cul_model_sel <- INLAModelSel_sc(resp, covar, Random = "site", RandomModel = "iid", Family = "gaussian",  Data = lm)

#Shows variables removed from all variables
f_covar <- inla_cul_model_sel$Removed[[length(inla_cul_model_sel$Removed)]]
f_covar
#New models using the reduced variable amount-------
#Create function combinations of all models with and without the grouping factor (f(farm, model = "iid"))

f2 = ibi ~ scale(turb) + scale(ph) + scale(cond) + scale(percent_dev) + scale(percent_ag) + scale(percent_dev_riparian) + scale(percent_dev_riparian) +
f(site, model = "iid")
#f3 = EHDV_Farm_ ~ scale() +

#full model with random
test2 <- inla(f2, family = "gaussian", control.compute = list(dic = TRUE), data = lm)
summary(test2)
Efxplot(list(test2))

#reduced model with random
test3 <- inla(f3, family = "gaussian", control.compute = list(dic = TRUE), data = l_m_nb_col)
summary(test3)
Efxplot(list(test3))


Model_list <- list(inla_cul_model, test2, test3)
sapply(Model_list, function(f) f$dic$dic)
#Compares dic values between the different models
INLADICFig(Model_list, ModelNames = c("Full_Ran", "Red_Ran", "Red_wo_Ran"))


```